---
title: "The Dynamics of Work Hours and Work-Life Balance: Insights from the General Social Survey"
author: 
  - Qin He
  - Shuyuan Zheng
  - Zijun Meng
thanks: "Code and data are available at: https://github.com/FrankMengZJ/work-hours"
date: today
date-format: long
abstract: "This paper examines the relationship between work hours and life quality using data from the US General Social Survey. Findings indicate that higher work satisfaction and financial comfort are associated with fewer work hours, highlighting inequalities based on social class. These insights underscore the importance of considering work hours in efforts to improve overall life satisfaction and work-life balance."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(here)
library(knitr)
library(kableExtra)

cleaned_data<-read_csv(file=here("data/analysis_data/analysis_data.csv"))

```


# Introduction

Work hours, as a central facet of our daily routines, have profound implications for our well-being and life satisfaction. The quest for an optimal work-life balance (WLB) has become more pressing in recent years, amid reports of increasing work hours and their consequent impact on individuals' quality of life worldwide. The concept of WLB, however, is multifaceted and varies in its operational definition across different studies, reflecting the complexity of balancing professional obligations with personal life in a way that promotes individual growth and well-being.

@WLB conceptualize WLB as "The individual’s perception that work and non-work activities are compatible and promote growth in accordance with an individual’s current life priorities," highlighting the subjective nature of achieving balance based on personal priorities and life stages. Similarly, the Canadian Department of Labor's perspective, as cited in @WFP, defines WLB more expansively as "a self-defined, self-determined state of well-being that a person can reach, or can set as a goal, that allows them to manage effectively multiple responsibilities at work, at home, and in their community; it supports physical, emotional, family, and community health, and does so without grief, stress or negative impact." These definitions underscore the personalized and dynamic nature of work-life balance, suggesting that achieving it involves managing various dimensions of life in a harmonious manner that minimizes stress and maximizes well-being.
Drawing from the rich data provided by the US General Social Survey (GSS) conducted by @data, this paper seeks to explore the intricate relationships between work hours and various factors, including financial satisfaction, personal well-being, social class perceptions, and demographic characteristics. The GSS dataset, renowned for its comprehensive coverage of the American population's attitudes and behaviors, offers a unique lens through which we can assess these dynamics over time.

The paper is structured as follows: The Data section details the sources, limitations, and preprocessing of the GSS dataset. The Results section presents our findings, showcasing the relationships between work hours and various factors across different demographic groups.

# Data
Data used in this paper are retrieved from the US General Social Survey (GSS) from NORC at @data. We retrieved demographic data as well as survey questions related to work hours from 1972 to 2022.

## Source Data
Source Data For demographic factors and questions related to work hours, we retrieved the following data as described in @tbl-data。
```{r}
#| label: tbl-data
#| tbl-cap: "Data Retrived"
#| echo: false


data <- data.frame(
  Variable = c("Year","id","age","degree","marital","babies","preteen","teens","adults","wrkstat","earnrs","income","relig","happy","health","satjob","satfin","class","hrs","sphrs"),
  Description = c("Calendar year of the survey", "Unique identifier for each respondent", "Respondent's age", "Highest degree received by the respondent", "Respondent's marital status", "Number of babies in the household", "Number of preteen children in the household", "Number of teenage children in the household", "Number of adult members in the household", "Work status of the respondent", "Number of earners in the household", "Total family income recorded in the year", "Religious affiliation of the respondent", "Respondent's overall happiness", "Respondent's self-reported health status", "Respondent's job satisfaction", "Respondent's financial satisfaction", "Self-identified social class of the respondent", "Total hours worked by the respondent in a week", "Total hours worked by the respondent's spouse in a week"),
  Example = c("2022", "12345", "56", "Bachelor's", "Divorced", "2", "1", "0", "2", "Full-time", "1", "$50,000 to $59,999", "3", "3", "3", "3", "4", "Middle class", "40", "35")
)
kable(data)
```
## Data Limitations
### Non-response Error
The transition to web surveys in 2021 resulted in a lower response rate (17%) compared to the traditionally higher rates achieved through in-person surveys. Although this rate is high for web surveys, it marks a significant departure from past GSS methodologies. Differential participation rates across years could influence the representativeness of the survey data. While efforts to adjust for these differences using post-stratification weighting are made, the potential for non-response bias impacting trend comparisons remains a concern.

### Remote Work
For remote work during the COVID-19 period, the shift to remote work can significantly affect the true total working hours reported. Without the clear boundaries that separate work from personal time in an office setting, remote employees might work longer hours, either by starting earlier, finishing later, or working intermittently across the day.

The transition to remote work has also introduced challenges in accurately reporting work hours. Employees working from home may have their work hours categorized as "NA" (not available) due to inconsistencies in self-reporting.

## Data Cleaning
Data was cleaned and analyzed using the open source statistically programming language @R, using functionalities from @tidyverse, and @haven.

The dataset was filtered to select key variables of interest, such as working hours, respondent demographics, and survey responses. This step involved selecting specific columns from the raw GSS data, including year, identification variables, work hours, age, educational attainment, marital status, number of children in different age groups, work status, income, and measures of subjective well-being among others.

Missing values for work hours (hrs1, hrs2) and spouse's work hours (sphrs1, sphrs2) were set to 0, indicating no reported hours. Similar treatment was applied to variables indicating the number of children in different age groups and the number of earners in the household. New variables were created to represent total work hours (hrs) and spouse's total work hours (sphrs) by summing the respective variables. Records with no reported work hours or earners were filtered out to focus on employed respondents. 


# Results

## Average working hours
@fig-year shows the fluctuating average working hours per week from 1980 to beyond 2020, with notable peaks suggesting periods of increased work hours and a recent sharp decline, possibly indicating economic or policy changes.

```{r}
#| label: fig-year
#| fig-cap: Average Working Hours per Year
#| echo: false

hrs_by_years<-
  group_by(cleaned_data, year)|>
  summarize(mean=mean(hrs))

ggplot(hrs_by_years,aes(x=year,y=mean))+
  geom_line() +
  geom_point() +
  theme_minimal() + 
  labs(x = "Year",
       y = "Average Working Hours per Week")
```

@fig-class displays the average working hours per week by social class,, the lower class works the most hours, and the upper class works the least, indicating a socioeconomic disparity in work hours.


```{r}
#| label: fig-class
#| fig-cap: Average Working Hours per Class
#| echo: false
#| warning: false
#| message: false

hrs_by_class <-
  cleaned_data |>
  filter(is.na(cleaned_data$class)==FALSE&cleaned_data$class!=5) |>
  group_by(class)|>
  summarize(mean=mean(hrs)) |>
  mutate(class_str=as.character(class))

hrs_by_class$class_str[hrs_by_class$class_str == '1'] <- 'Lower class'
hrs_by_class$class_str[hrs_by_class$class_str == '2'] <- 'Working class'
hrs_by_class$class_str[hrs_by_class$class_str == '3'] <- 'Middle class'
hrs_by_class$class_str[hrs_by_class$class_str == '4'] <- 'Upper class'

ggplot(hrs_by_class, aes(x = factor(class_str,level = c('Lower class','Working class','Middle class','Upper class')), y = mean, fill = class_str)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Average Working Hours per Class", x = "Class", y = "Average Working Hours per Week", fill ="Class")
```

@fig-age and @fig-minors displayed average working hours across different age groups. @fig-age presents a typical lifecycle pattern of average working hours per week in relation to age, showing an increase through early adulthood to middle age, then a decline as individuals pass the age of 60, likely due to retirement. 


```{r}
#| label: fig-age
#| fig-cap: Average Working Hours per Age per Week
#| echo: false
#| warning: false
#| message: false
hrs_by_age<-
  group_by(cleaned_data, age)|>
  summarize(mean=mean(hrs))

ggplot(hrs_by_age,aes(x=age,y=mean))+
  geom_line() +
  geom_point() +
  theme_minimal() + 
  labs(title = "Average Working Hours per Age",
       x = "age",
       y = "Average Working Hours per Week")

```



```{r}
#| label: fig-minors
#| fig-cap: Average Working Hours per Number of Minors
#| echo: false
#| warning: false
#| message: false
hrs_by_num_of_minors <-
  cleaned_data |>
  mutate(minors = babies + preteen + teens)|>
  group_by(minors)|>
  summarize(mean = mean(hrs))

ggplot(hrs_by_num_of_minors,aes(x=minors,y=mean))+
  geom_line() +
  geom_point() +
  theme_minimal() + 
  labs(x = "Number of Minors",
       y = "Average Working Hours per Week")

```


@fig-ratio reveals a high degree of variability in working hours when the ratio of family members to earners is low, suggesting that in households where the number of earners is relatively high compared to family members, the working hours are inconsistent. This could be due to different work patterns or job types among earners.

As the ratio increases, which implies fewer earners compared to family members, the average working hours per week seem to decrease and stabilize. Interestingly, after the ratio surpasses 5, there is a significant drop in variability and a general downward trend in average working hours. This suggests that in households with a higher number of dependents per earner, the average working hours are lower and more consistent, possibly due to the constraints of balancing work and caregiving responsibilities.

```{r}
#| label: fig-ratio
#| fig-cap: Average Working Hours per Ratio
#| echo: false
#| warning: false
#| message: false
hrs_by_ratio <-
  cleaned_data |>
  mutate(ratio = (babies + preteen + teens + adults)/earnrs) |>
  group_by(ratio) |>
  summarize(mean = mean(hrs))

ggplot(hrs_by_ratio,aes(x=ratio,y=mean))+
  geom_line() +
  geom_point() +
  theme_minimal() + 
  labs(x = "Ratio of Family Members and Earners",
       y = "Average Working Hours per Week")
```

## Work Satisfaction

@fig-satjob illustrates a clear downward trend in average working hours as work satisfaction increases, suggesting a potential relationship between these two variables. This indicates that employees with higher work satisfaction tend to have fewer average working hours, which could be interpreted as a sign that better work-life balance leads to higher job satisfaction, or that happier employees are more efficient and thus need less time to complete their work. However, it’s crucial to note that correlation does not equal causation, and further analysis would be required to understand the underlying reasons for this relationship.


```{r}
#| label: fig-satjob
#| fig-cap: Work Satifsaction and Average Working Hours
#| echo: false
#| warning: false
#| message: false
hrs_by_satjob<-
  group_by(cleaned_data, satjob )|>
  summarize(mean=mean(hrs))

ggplot(hrs_by_satjob,aes(x=satjob ,y=mean))+
  geom_line() +
  geom_point() +
  theme_minimal() + 
  labs(x = "Work Satifsaction",
       y = "Average Working Hours per Week")

```

In @fig-satjob_by_hrs, the distribution of points indicates that for the majority of the working hours, average work satisfaction remains relatively stable, hovering around the 1.5 mark. There's no clear trend or correlation observable for most of the data range; however, there is a noticeable spike in work satisfaction at the highest end of the working hours spectrum. This could represent an anomaly or a specific situation where individuals working exceptionally long hours are reporting higher satisfaction, possibly due to high compensation, passion for their work, or other job-related benefits.


```{r}
#| label: fig-satjob_by_hrs
#| fig-cap: Working Hours and Average Work Satifsaction
#| echo: false
#| warning: false
#| message: false
satjob_by_hrs<-
  cleaned_data |>
  filter(is.na(cleaned_data$satjob)==FALSE)|>
  group_by(hrs )|>
  summarize(mean=mean(satjob))

ggplot(satjob_by_hrs,aes(x=hrs,y=mean))+
  geom_line() +
  geom_point() +
  theme_minimal() + 
  labs(x = "Working Hours per Week",
       y = "Average Work Satifsaction") +
  expand_limits(y=0)
```

## Financial satisfaction

In @fig-satfin, it exhibits a downward slope, indicating that as financial satisfaction increases, the average working hours decrease. This could suggest that individuals who are more financially satisfied may be working fewer hours, perhaps due to higher wages or the ability to achieve a comfortable living without needing to work longer hours. Conversely, those with lower financial satisfaction may be working more hours, potentially due to lower wages requiring more work to meet financial needs.

```{r}
#| label: fig-satfin
#| fig-cap: Financial Satifsaction and Average Working Hours
#| echo: false
#| warning: false
#| message: false
hrs_by_satfin<-
  group_by(cleaned_data, satfin )|>
  summarize(mean=mean(hrs))

ggplot(hrs_by_satfin,aes(x=satfin ,y=mean))+
  geom_line() +
  geom_point() +
  theme_minimal() + 
  labs(x = "Financial Satifsaction",
       y = "Average Working Hours per Week")
```





# Discussion

## Work Hours and Socio-Economic Disparities

The data indicate a nuanced relationship between work hours and social class, with @fig-class illustrating that individuals from lower socio-economic backgrounds tend to work longer hours compared to their upper-class counterparts. This trend suggests a reinforcement of existing socio-economic disparities, where those in lower social classes may engage in longer work hours out of necessity rather than choice, potentially due to lower wage rates or the type of employment available to them. Conversely, the upper class, possibly benefiting from higher wages or salaries, has the luxury of working fewer hours, highlighting a socio-economic advantage that allows for a better work-life balance. Overall, this situation will continue to widen economic inequality and will continue to create significant differences in the quality of life. People from lower socio-economic backgrounds will continue to lose leisure time and opportunities for personal development, creating a vicious circle.

## Lifecycle Patterns and the Impact of Age on Work Hours

Our findings, shown in @fig-age, present a life-cycle pattern of hours worked, reflecting a peak in middle age followed by a decline after age 60 (possibly due to retirement). Specifically, the golden years of adulthood are largely devoted to labor, which often tends to come at the expense of personal development and leisure. While work experience increases with age, the decline in hours worked after age 60 suggests a social acceptance of retirement or consideration of the physical condition of older people and allows them to shift their priorities to leisure, health and family. However, this life-cycle pattern also reflects the challenges faced by different age groups, in particular the need to balance the pressures of work, family responsibilities and personal well-being for middle-aged people, many of whom need to devote a significant amount of time to work while the cost of unemployment can be difficult for them to bear, and for older people, the need for the labor market to balance the risks to their health and the industry benefits of their wealth of work experience. The advantages of older workers, especially those with experience in education and research, are irreplaceable by younger workers.


## Better Financial Satisfaction with Work Hours

As shown in @fig-satfin, the inverse relationship between financial satisfaction and hours worked provides insights into the dynamics of work and finances. It suggests that those who are more financially satisfied tend to work fewer hours, possibly because they either earn enough money to satisfy their needs in fewer hours of work or because they value time over money and choose to work less and possibly live more. On the other hand, those with lower financial satisfaction may find themselves working longer hours either to break even or to pursue financial goals that are difficult to achieve due to the wage gap or cost of living. This finding challenges the conventional wisdom that more work equals more financial security. While it is true that different types of workers produce different amounts of value per unit of time, blindly constraining those who are highly specialized to do more work based on the notion of "the more you can do, the more you have to do" does not lead to greater financial satisfaction even if they are paid more. With the current development of human productivity, the labor market can gradually provide a more professional threshold that is not so high work, to let as many people as possible participate in the work which can reduce the intensity of the work of some highly professional talent, also as far as possible to prevent the culture of involution so that society can shift to a model of work that can improve economic satisfaction and do not require too many hours of work. Everyone has a job, everyone has spending power, but not too much work time for one person, creating a virtuous socio-economic cycle. While many companies are already doing this, there is still a long way to go before the idealized labor market is completely accomplished.

# Weaknesses and Next Steps

This study’s reliance on self-reported data and the transition to web surveys introduces potential biases and comparability issues across the dataset. Moreover, it doesn't fully explore the qualitative dimensions of work, such as job satisfaction and workplace culture, or delve into regional and sector-specific variations that might affect work-life balance.

Future Directions:

1. Longitudinal Research: To better understand the long-term impact of work hours on well-being, future studies should adopt longitudinal approaches.
2. Qualitative Insights: Incorporating qualitative research can deepen our understanding of the subjective experience of work and its balance with personal life.
3. Cross-Cultural and Sector-Specific Studies: Exploring differences across cultures and industry sectors can offer nuanced insights into work-life balance challenges and solutions.
4. Policy Analysis: Evaluating the effectiveness of work-related policies can guide efforts to improve work-life balance and reduce social disparities.

# Conclusion

In conclusion, this paper has navigated through the complex terrain of work hours and their implications on various aspects of life, shedding light on the multifaceted relationship between work-life balance (WLB) and individual well-being. By leveraging the extensive data provided by the US General Social Survey (GSS), we have been able to discern patterns and relationships that underscore the significance of achieving a harmonious balance between professional responsibilities and personal life. Our findings reveal that work hours indeed have a considerable impact on financial satisfaction, personal well-being, perceptions of social class, and demographic characteristics, illustrating the diverse ways in which work-life balance affects individuals across different walks of life.



\newpage

\appendix

# Appendix {-}
Link to the survey: https://forms.gle/Bm6JqTnuFAyVWwMq8




\newpage


# References


